A Standardized Effect Size for Evaluating the Comparing the Strength of Phylogenetic Signal

### 2020-07-15 Manuscript/OldVsns/Manuscript-7-MLC.Rmd
pnas vsn

Several analytical tools have been developed to quantify phylogenetic signal in phenotypic datasets, including measures of serial independence [$\mathbf{C}$: @Abouheif1999], autocorrelation estimates [$I$: @Gittleman1990], statistical ratios of trait variation relative to what is expected given the phylogeny [$\kappa$: @Blomberg_et_al2003; @Adams2014a], and scaling parameters used in maximum likelihood fitting of the data to the phylogeny [$\lambda$: @Pagel1999], among others [e.g., @Klingenberg2010]. The statistical properties of these methods -- namely type I error rates and power -- have also been investigated to determine when phylogenetic signal can be detected and under what conditions [e.g., @Munkemuller_et_al2012; @Pavoine2012;  @DinizFilho2012; @Adams2014a; @MolinaVenegas2017; see also @Revell_et_al2008; @Revell2010]. One of the most widely used methods for characterizing phylogenetic signal in macroevolutionary studies is Pagel's $\lambda$ [@Pagel1999]. The ~~($\lambda$)~~ parameter ($\lambda$) transforms the lengths of the internal branches of the phylogeny to improve ~~a~~ THE fit of data to the phylogeny via maximum likelihood [@Pagel1999; @Freckleton_et_al2002]. Pagel's $\lambda$ ranges from $0\to1$, with larger values signifying a greater dependence of observed trait variation on the phylogeny. Pagel's $\lambda$ also has the appeal that it may be included in phylogenetic generalized least-squares regression (PGLS) to account for the degree of phylogenetic signal in comparative analyses [see @Freckleton_et_al2002]. \hfill\break

In addition to functioning as a parameter that is tuned for appropriate analysis, $\lambda$ can function as a descriptive statistic ~~to describe~~ CHARACTERIZING the relative strength of phylogenetic signal in phenotypic traits to determine the extent to which shared evolutionary history has influenced trait covariation among taxa. ~~The appeal~~ THIS DESCRIPTIVE UTILITY of $\lambda$ ~~as a descriptive statistic for evolutionary biologists~~ is ~~a~~ THE basis for interpreting "weak" versus "strong" phylogenetic signal~~,~~ (i.e., small versus large values of $\lambda$, respectively)~~,~~ in a comparative sense [e.g., @DeMeester2019; @Pintanel2019;@Su2019]. Indeed, statements regarding the strength of phylogenetic signal based on $\lambda$ are rather common in the evolutionary literature. For instance, of the 204 papers published in 2019 that estimated and reported Pagel's $\lambda$ (found from a literature survey we conducted in Google.scholar), 40% EXPLICITLY interpreted the strength of phylogenetic signal for at least one phenotypic trait. Further, because nearly half of the 1,572 $\lambda$ values reported were near 0 or 1 (Figure 1) where the biological interpretation of $\lambda$ is known, this percentage is even higher. \hfill\break

@@

Various other approaches use $\lambda$ as a parameter that can be ~~varied~~ ADJUSTED for inferences akin to sensitivity analysis. For instance, some have performed likelihood ratio tests that compare observed model fits to those obtained when $\lambda=0$ or $\lambda=1$ [@Freckleton_et_al2002; @Cooper2010; @Bose2019] or evaluated whether observed $\lambda$ differs from an expected $\lambda$, based on confidence intervals generated for the expected value [@Vandelook2019].  Qualitative comparisons of $\lambda$ estimates have also been performed for multiple traits on the same phylogenetic tree to infer whether the strength of phylogenetic signal is greater in one trait as compared to another [e.g., @Liu2019; @Bai2019]. \hfill\break

It seems intuitive to interpret the strength of phylogenetic signal based on the value of $\lambda$, as $\lambda$ is a parameter on a bounded scale ($0\to1$) for which interpretation of its extremal points are understood. Specifically, $\lambda=0$ represents no phylogenetic signal, while $\lambda=1$ is phylogenetic signal as expected under Brownian motion. However, equating values of $\lambda$ directly to the strength of phylogenetic signal presumes two important statistical properties that have not been fully explored. First, it presumes that values of $\lambda$ can be precisely estimated, as biological inferences regarding the strength of phylogenetic signal depend on high accuracy in its estimation. Therefore, understanding the precision in estimating $\lambda$ is paramount. One study [@Boettiger_et_al2012] found that estimates of Pagel's $\lambda$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation it was concluded that insufficient data (i.e., ~~the~~ LOW number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise at greater sample sizes [@Cohen1988]. However, this conclusion also implies that the precision of $\lambda$ remains constant across its range ($\lambda = 0 \to 1$); an assumption that to date, has not been verified. Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we lack a general understanding of the precision with which $\lambda$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while estimates of $\lambda$ are within a bounded scale ($0\to1$), this does not *de-facto* imply that the estimated values of this parameter correspond to the actual strength of the underlying input signal in the data. For this to be the case, $\lambda$ must be a statistical effect size. Effect sizes are a measure of the magnitude of a statistical effect in data, represented on a common scale [@Glass1976; @Cohen1988]. Effect sizes have widespread use in many areas of the quantiative sciences, as they represent measures that may be readily summarized across datasets as in meta-analyses [@HedgesOlkin1985; @Glass1976; @Arnqvist1995], or compared among datasets [e.g., @AdamsCollyer2016; @AdamsCollyer2019b]. Unfortunatley, not all model parameters and descriptive statistics are effect sizes, and thus many summary measures must first be converted to statistics with standardized units (i.e., conversion to an effect size) for meaningful comparison [see @Rosenthal1994]. As a consequence, it follows that only if $\lambda$ is a statistical effect size can comparisons of estimates across datasets be interpretable. ~~For the case of $\lambda$, this~~ HOWEVER, THE CALCULATION AND STATISTICAL BEHAVIOR OF $\lambda$ AS AN EFFECT SIZE has not yet been explored.  \hfill\break

In this study, we evaluate the precision of Pagel's $\lambda$ for estimating known levels of phylogenetic signal in phenotypic data. We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. We find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that the precision in estimating $\lambda$ is not constant across its range. Rather, there is decreased precision when input levels of phylogenetic signal are of intermediate strength. Additionally, the same estimated values of $\lambda$ may be obtained from datasets containing vastly different input levels of phylogenetic signal. Thus, $\lambda$ is not a reliable indicator of the strength of phylogenetic signal in phenotypic data. We then describe a standardized effect size for measuring the strength of phylogenetic signal in phenotypic datasets and apply the concept to two common measures of phylogenetic signal: $\lambda$ and $\kappa$. Through simulations we find that the precision of effect sizes based on $\lambda$ ($Z_{\lambda}$) are less reliable than that those based on $\kappa$ ($Z_\kappa$), implying that $Z_\kappa$ is a more robust effect size measure. We also propose a two-sample test statistic that may be used to compare the strength of phylogenetic signal among datasets, and provide an empirical example to demonstrate its use. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often inaccurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $\kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

@@

## *The Precision of $\lambda$ is Variable*

We conducted a series of computer simulations to evaluate the precision of Pagel's $\lambda$. Our primary simulations were based on pure-birth phylogenies; however, we also evaluated patterns on both balanced and pectinate trees to determine whether tree shape affected our findings (see Supporting Information). First we generated 50 pure-birth phylogenies at each of six different tree sizes, ranging from 32 to 1024 taxa ($n=2^5 - 2^{10}$). Next, we rescaled the simulated phylogenies by multiplying the internal branches by $\lambda_{in}$, using 21 intervals of 0.05 units across its range ($\lambda_{in} = 0.0 \to 1.0$), resulting in 1050 scaled phylogenies at each level of species richness ($n$). Continuous traits were then simulated on each phylogeny under a Brownian motion model of evolution to obtain datasets with differing levels of phylogenetic signal, that ranged from no phylogenetic signal (when $\lambda_{in} =0$), to phylogenetic signal reflecting Brownian motion (when $\lambda_{in} =1$). For each dataset we then estimated phylogenetic signal ($\lambda_{est}$), and calculated the variance of $\lambda$ ($\sigma^2_\lambda$) across datasets at each input level of phylogenetic signal and level of species richness as an estimate of precision. We verified that the variance of traits simulated had no effect on phylogenetic signal estimation. \hfill\break

We also evaluated the precision of $\lambda$ when estimated in PGLS regression and ANOVA (i.e., $Y\sim{X}$). Here, an independent variable $X$ was simulated on each rescaled phylogeny under a Brownian motion model of evolution (for PGLS regression). For phylogenetic ANOVA, random groups ($X$) were obtained by simulating a discrete (binary, 0 or 1) character on each phylogeny. Next, the dependent variable was simulated in such a manner as to contain a known relationship with $X$ plus random error containing phylogenetic signal. This was accomplished as: $Y=\beta{X}+\epsilon$.  The association between $Y$ and $X$ was modeled using a range of values: $\beta=(0.0,0.25, 0.5, 0.75,1.0)$, and the residual error ($\epsilon$) was modeled to contain phylogenetic signal simulated under a Brownian motion model of evolution on each rescaled phylogeny: $\epsilon=\mathcal{N}(\mu=0,\sigma=\sigma^2\mathbf{C})$: [see @Revell2010 for a similar simulation design]. The fit of the phylogenetic regression was estimated using maximum likelihood, and parameter estimates ($\beta_{est}$ and $\lambda_{est}$) were obtained. We then calculated precision estimates ($\sigma^2_\lambda$) at each input level of phylogenetic signal and level of species richness.  We verified that the amount of residual variance simulated had no effect on $\sigma^2_\lambda$ but did influence the precision of coefficients estimated from the linear model (precision increased with smaller $\epsilon$, as expected).  \hfill\break

@@

*Results.* We found that the precision of $\lambda_{est}$ varied widely across simulation conditions. Predictably, precision improved as the number of species increased (Figure 2). This confirmed earlier findings of Boettiger et al. [-@Boettiger_et_al2012], and adhered to parametric statistical theory. However, in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: $\lambda_{est}= 0.0\to 0.985$), revealing that estimates of $\lambda$ were not a reliable indicator of input phylogenetic signal. Importantly, the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal. The worst precision was observed at intermediate levels of phylogenetic signal ($\lambda_{in}\approx0.5$), while precision improved as input levels approached the extremes of $\lambda$'s range (i.e., $\lambda_{in}\rightarrow0$ & $\lambda_{in}\rightarrow1$). Thus, estimates of $\lambda$ were least reflective of the true input signal at intermediate values. Additionally, even at large levels of species richness, we found that the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; $\lambda_{in} = 0.25 \to 0.65$). These findings were particularly unsettling when considered in light of our literature survey. Over one quarter of the $\lambda$ estimates published in empirical studies (421 of 1,572) were between $\lambda=0.25$ and $\lambda=0.75$ (Figure 1). This range reflected the region that our simulations identified as being the least reliable in terms of accurately characterizing levels of phylogenetic signal, yet 30% of these mid-range empirical estimates were explicitly interpreted in terms of the strength of phylogenetic signal that they represented (i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

Finally, when $\lambda$ was co-estimated with regression parameters in PGLS regression and ANOVA, the results of our simulations were quite similar. Regression parameters ($\beta$) were accurately estimated, confirming earlier findings of Revell [-@Revell2010] (see Supporting Information). However, estimates of phylogenetic signal ($\lambda$) were less precise (Figure 3; see also Supporting Information), and the spread of $\lambda_{est}$ was similar to that observed when $\lambda$ was estimated for only the dependent variable, as in Figure 2. Taken together, these findings reveal that $\lambda_{est}$ does not precisely characterize ~~observed~~ KNOWN levels of phylogenetic signal in phenotypic datasets, and that biological interpretations of the strength of phylogenetic signal based on $\lambda$ may be highly inaccurate. \hfill\break

@@

An effect size based on $\lambda$ could be envisioned, which is found as:

\begin{align}
    Z_{\lambda}=\frac{\lambda_{obs}-0}{\hat\sigma_\lambda}.
\end{align}

@@

To evaluate the utility of $Z_\kappa$ and $Z_\lambda$ we calculated both effect sizes for the simulated datasets generated above, and summarized the precision of each using its variance ($\sigma^2_{Z_\kappa}$ and $\sigma^2_{Z_\lambda}$, Figure 4: additional results in the Supporting Information). Here two things are evident. First, estimates of $Z_\kappa$ linearly track the input phylogenetic signal whereas estimates of $Z_\lambda$ do not (Figure 4A, B). Thus, actual changes in the strength of phylogenetic signal are reflected more evenly in the corresponding values of the effect size $Z_\kappa$. Second, the precision of $Z_\kappa$ is considerably more stable as compared with $Z_\lambda$. This may be seen by calculating the coefficients of variation for the set of precision estimates (i.e., $\sigma^2_{Z_\kappa}$ and $\sigma^2_{Z_\lambda}$) across input levels of phylogenetic signal.  Coefficients of variation in the precision of $Z_\kappa$ were up to an order of magnitude smaller ~~for~~ than for $Z_\lambda$ (Figure 4C), implying that estimates of the strength of phylogenetic signal were more reliable and robust when using $Z_\kappa$. \hfill\break

@@

To demonstrate the utility of $\hat{Z}_{12}$ we quantified and compared the strength of phylogenetic signal of two phenotypic traits across species of plethodontid salamander. The data were part of a series of studies examining macroevolutionary trends in phenotypic diversification in this group [@Baken2019; @Baken2020]. Our dataset contained surface area to volume ratios (SA:V) and relative body width ($\frac{BW}{SVL}$) for 305 species (Figure 5A). For SA:V, 11 linear body measurements from 2,781 individuals were taken, from which estimates of the surface area and volume of the head, body, and tail were calculated and subsequently combined to arrive at the SA:V for each indiviual [for mathematical details see @Baken2020]. Species means were then obtained. Likewise, body size (SVL) and body width (BW) measurements were taken from 3,371 individuals, and species means of relative body width ($\frac{BW}{SVL}$) were calculated [data from @Baken2019]. A time-dated molecular phylogeny for the group [@Bonett2017] was then pruned to match the species in the PHENOTYPIC dataset. ~~, resulting in a phylogeny and corresponding phenotypic dataset containing 305 species.~~ <!-- number of species already mentioned earlier in this paragraph--> The phylogenetic signal in each trait was then characterized using $\kappa$, which was converted to its effect size ($Z_\kappa$) using `geomorph` 3.2.1 [@AdamsOtarola2013; @AdamsGeomorph]. Finally, the strength of phylogenetic signal was compared across traits using $\hat{Z}_{12}$ as described above (**to be incorporated in `geomorph` upon manuscript acceptance**). \hfill\break

>>>>

Several analytical tools have been developed to quantify phylogenetic signal in phenotypic datasets, including measures of serial independence [$\mathbf{C}$: @Abouheif1999], autocorrelation estimates [$I$: @Gittleman1990], statistical ratios of trait variation relative to what is expected given the phylogeny [$\kappa$: @Blomberg_et_al2003; @Adams2014a], and scaling parameters used in maximum likelihood fitting of the data to the phylogeny [$\lambda$: @Pagel1999], among others [e.g., @Klingenberg2010]. The statistical properties of these methods -- namely type I error rates and power -- have also been investigated to determine when phylogenetic signal can be detected and under what conditions [e.g., @Munkemuller_et_al2012; @Pavoine2012;  @DinizFilho2012; @Adams2014a; @MolinaVenegas2017; see also @Revell_et_al2008; @Revell2010]. One of the most widely used methods for characterizing phylogenetic signal in macroevolutionary studies is Pagel's $\lambda$ [@Pagel1999]. The parameter ($\lambda$) transforms the lengths of the internal branches of the phylogeny to improve the fit of data to the phylogeny via maximum likelihood [@Pagel1999; @Freckleton_et_al2002]. Pagel's $\lambda$ ranges from $0\to1$, with larger values signifying a greater dependence of observed trait variation on the phylogeny. Pagel's $\lambda$ also has the appeal that it may be included in phylogenetic generalized least-squares regression (PGLS) to account for the degree of phylogenetic signal in comparative analyses [see @Freckleton_et_al2002]. \hfill\break

In PGLS analysis, $\lambda$ is a parameter that is tuned via log-likelihood profiling for its estimation.  After obtaining an optimized $\lambda$ value, $\hat{\lambda}$, likelihood ratio tests can be performed to compare observed model fits using $\hat{\lambda}$ to those obtained when $\lambda=0$ or $\lambda=1$ to infer whether phylogenetic signal differs from no signal or a Brownian motion (BM) model of evolutionary divergence, respectively [@Freckleton_et_al2002; @Cooper2010; @Bose2019].  Confidence limits on $\hat{\lambda}$ can also be estimated based on percentiles of a $\chi^2$ distribution and where these values are mapped onto a log-likelihood profile [*sensu* @Freckleton_et_al2002].  With respect to confidence limits, one can ascertain whether $\lambda=0$ or $\lambda=1$ is contained within an interval of $\hat{\lambda}$, as an analog to likelihood ratio tests [see, e.g., @Vandelook2019].  It is, therefore, tempting to regard $\hat{\lambda}$ as a descriptive statistic that measures the relative strength of phylogenetic signal, for inferring the extent to which shared evolutionary history has influenced trait covariation among taxa.  The appeal of $\hat{\lambda}$ as a descriptive statistic is that it serves as the basis for interpreting "weak" versus "strong" phylogenetic signal; i.e., small versus large values of $\hat{\lambda}$, respectively, in a comparative sense [e.g., @DeMeester2019; @Pintanel2019;@Su2019]. For example, qualitative comparisons of $\hat{\lambda}$ for multiple traits on the same phylogenetic tree provide a basis for inferring whether the strength of phylogenetic signal is greater in one trait as compared to another [e.g., @Liu2019; @Bai2019].  Indeed, statements regarding the strength of phylogenetic signal based on $\hat{\lambda}$ are rather common in the evolutionary literature. For instance, of the 204 papers published in 2019 that estimated and reported Pagel's $\lambda$ (found from a literature survey we conducted in Google.scholar), 40% explicitly interpreted the strength of phylogenetic signal for at least one phenotypic trait. Further, because nearly half of the 1,572 $\hat{\lambda}$ values reported were near 0 or 1 (Figure 1) where the biological interpretation of $\lambda$ is known ($\lambda=0$ represents no phylogenetic signal, while $\lambda=1$ is phylogenetic signal as expected under BM), this percentage is even higher. \hfill\break

@@

# As a statistic that measures phylogenetic signal, how should $\hat{\lambda}$ behave?

It seems intuitive to interpret the strength of phylogenetic signal based on the value of $\hat{\lambda}$, as $\lambda$ is a parameter on a bounded scale ($0\to1$) for which interpretation of its extremal points are understood. However, equating values of $\hat{\lambda}$ directly to the strength of phylogenetic signal presumes two important statistical properties that have not been fully explored. First, it presumes that values of $\hat{\lambda}$ can be precisely estimated, as biological inferences regarding the strength of phylogenetic signal depend on high accuracy in its estimation. (In order to be precise, the log-likelihood profile would have to be peaked rather than flat.)  Therefore, understanding the precision in estimating $\hat{\lambda}$ is paramount. As a prortional random variable bounded by 0 and 1, we might expect that $\hat{\lambda}$ follows A Bernoulli distribution; i.e., branch lengths in a tree are scaled proportionally to the probability that data arise from a BM process.  As such, we would also expect in simulation experiment that given a known $\lambda$ value used to generate random data on a tree: the mean of an empirical sampling distribution of $\hat{\lambda}$ would approximately equal $\lambda$; the standard error of $\hat{\lambda}$ would be largest at intermediate values of $\lambda$, and predictable over the range of $\lambda$ with respect to treesize; the distribution of $\hat{\lambda}$ would be symmetric at intermediate values of $\lambda$ and more skewed toward values of 0 or 1; and that the distribution of $\hat{\lambda}$ will be more platykurtic at intermediate values of $\lambda$, becoming more leptokurtic toward 0 and 1.  These properties seemed to be superficially reasonable from the simulations perfomed by Münkemüller et al.[-@Munkemuller_et_al2012, see Fig. 2].  Their study simulated strength of phylogenetic signal as a varied weighted-average of data simulated on trees with $\lambda=0$ and $\lambda=1$, rather than from prescribed values of $\lambda$.  Nonetheless, their results revealed that as a random variable, $\hat{\lambda}$ resembled a Bernoulli random variable based superficially on statistical moments (mean, variance, skewness, and kurtosis; see their Fig. 2) with respect to strength of phylogenetic signal, for a given tree size.  However, for comparisons among tree sizes for a given strength of phylogenetic signal, median values of $\hat{\lambda}$ were slightly, positively associated with the number of species in a tree, especially at weaker strength of signals, suggesting $\hat{\lambda}$ might not be an accurate statistic across a range of tree sizes.   \hfill\break

Another study [@Boettiger_et_al2012] found that $\hat{\lambda}$ displayed less variation (i.e., greater precision) when data were simulated on a large phylogeny ($N=281$) as compared to a small one ($N=13$). From this observation it was concluded that insufficient data (i.e., low number of species) was the underlying cause of the increased variation across parameter estimates [@Boettiger_et_al2012]. Indeed, such a pattern is common with statistical estimators, as summary statistics and parameters are often more precise at greater sample sizes [@Cohen1988]. However, this conclusion could imply that the shape of log-likelihood profiles are assumed consistent with species number.  Such an assumption and the importance of tree shape have to date not been verified.  Thus, despite widespread use of Pagel's [-@Pagel1999] $\lambda$ in macroevolutionary studies, at present, we lack a general understanding of the precision with which $\hat{\lambda}$ can estimate levels of phylogenetic signal in phenotypic datasets. \hfill\break

Second, while $\lambda$ varies over a bounded scale ($0\to1$), this does not *de-facto* imply that the estimated values, $\hat{\lambda}$, of this parameter correspond to the actual strength of the underlying input signal in the data. For this to be the case, $\hat{\lambda}$ must be a statistical effect size. Effect sizes are a measure of the magnitude of a statistical effect in data, represented on a common scale [@Glass1976; @Cohen1988]. Effect sizes have widespread use in many areas of the quantiative sciences, as they represent measures that may be readily summarized across datasets as in meta-analyses [@HedgesOlkin1985; @Glass1976; @Arnqvist1995], or compared among datasets [e.g., @AdamsCollyer2016; @AdamsCollyer2019b]. Unfortunately, not all model parameters and descriptive statistics are effect sizes, directly, and thus many summary measures must first be converted to statistics with standardized units (i.e., conversion to an effect size) for meaningful comparison [see @Rosenthal1994]. As a consequence, it follows that only if $\hat{\lambda}$ can be cobverted to a statistical effect size can comparisons of estimates across datasets be interpretable. However, the calculation and statistical behavior of $\hat{\lambda}$ as an effect size has not yet been explored.  \hfill\break

In this study, we evaluate the precision of $\hat{\lambda}$ as an optimized parameter, used as a statistic, for estimating known levels of phylogenetic signal in phenotypic data (data simulated with known $\lambda$). We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of $\lambda$, to explore the degree to which $\hat{\lambda}$ correctly identifies known levels of phylogenetic signal, $\lambda$, and under which circumstances. We find that estimates of $\hat{\lambda}$ vary widely for a given input value of $\lambda$, and that the precision of $\hat{\lambda}$ is not constant across its range. Rather, there is decreased precision when input levels of phylogenetic signal are of intermediate strength. Additionally, the same values of $\hat{\lambda}$ may be obtained from datasets containing vastly different input levels of $\lambda$. Thus, $\hat{\lambda}$ is not a reliable indicator of the strength of phylogenetic signal in phenotypic data. We then describe a standardized effect size for measuring the strength of phylogenetic signal in phenotypic datasets and apply the concept to two common measures of phylogenetic signal: $\hat{\lambda}$ and $\kappa$. Through simulations we find that the precision of effect sizes based on $\hat{\lambda}$ ($Z_{\lambda}$) are less reliable than that those based on $\kappa$ ($Z_\kappa$), implying that $Z_\kappa$ is a more robust effect size measure. We also propose a two-sample test statistic that may be used to compare the strength of phylogenetic signal among datasets, and provide an empirical example to demonstrate its use. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ (as a descriptive statistic, not as a PGLS parameter) are often inaccurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $\kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

@@

## *The accuracy and precision of $\hat{\lambda}$ is Variable*

We conducted a series of computer simulations to evaluate the precision of $\hat{\lambda}$ based on varied input of Pagel's $\lambda$ to generate data. Our primary simulations were based on pure-birth phylogenies; however, we also evaluated patterns on both balanced and pectinate trees to determine whether tree shape affected our findings (see Supporting Information). First we generated 50 pure-birth phylogenies at each of six different tree sizes, ranging from 32 to 1024 taxa ($n=2^5 - 2^{10}$). For each simulation run, we rescaled the simulated phylogenies by multiplying the internal branches by $\lambda_{in}$, using 21 intervals of 0.05 units across its range ($\lambda_{in} = 0.0 \to 1.0$), resulting in 1050 scaled phylogenies at each level of species richness ($n$). Continuous traits were then simulated on each phylogeny under a BM model of evolution to obtain datasets with differing levels of phylogenetic signal, that ranged from no phylogenetic signal (when $\lambda_{in} =0$), to phylogenetic signal reflecting a BM model of evolution (when $\lambda_{in} =1$). For each dataset we then estimated phylogenetic signal ($\hat{\lambda}$), and calculated the standard deviation of ${\hat{\lambda}$ ($\sigma_{\hat{\lambda}}$) at each input level of phylogenetic signal and level of species richness as an estimate of precision. Random data samples were generated from a normal distribution, i.e.,  $\mathcal{N}(\mu=0,\sigma=\sigma^2\mathbf{C})$.  We verified that the level of $\sigma^2$ had no effect on $\hat{\lambda}$ or $\sigma^2_{\hat{\lambda}}$, but the results we present used $\sigma^2=1$. \hfill\break

We also evaluated the precision of $\hat{\lambda}$ when estimated in PGLS regression and ANOVA (i.e., $Y\sim{X}$). For these simulations, $Y=\beta{X}+\epsilon$ was used to generate data with varied phylogenetic signal and variable associations [*sensu* @Revell2010].  In the case of PGLS regression, $X$ was a variable simulated under a BM model of evolution, constrained by rescaling phylogeny according to $\lambda$, and in the case of ANOVA, $X$ was a balanced binary variable (values of 0 and 1), randomly ordered. The strength of variable associations ranged as $\beta=0.0,0.25, 0.5, 0.75$ and $1.0$. Residuals, $\epsilon$, were simulated under a BM model of evolution, constrained by rescaling phylogeny according to $\lambda$.  Simulated independent variables or residuals were generated from a normal distribution, i.e.,  $\mathcal{N}(\mu=0,\sigma=\sigma^2\mathbf{C})$, with the phylogenetic covariance matrix, $\mathbf{C}$, implicitly varied for each tree by the level of $\lambda$.  The fit of the phylogenetic regression was estimated using maximum likelihood, and parameter estimates ($\hat{\lambda}$ and $\hat{\beta}$) were obtained. We then calculated precision estimates ($\sigma^2_{\hat{\lambda}}$) at each input level of phylogenetic signal and level of species richness.  We verified that the level of $\sigma^2$ had no effect on $\sigma^2_{\hat{\lambda}}$ but did influence the precision of $\hat{\beta}$ estimated from the linear model (precision increased with smaller $\epsilon$, as expected). The results we present used $\sigma^2=1$.  \hfill\break

@@

*Results.*

superifical Bernoulli traits
innacuracy
precision

We found that the precision of $\lambda_{est}$ varied widely across simulation conditions. Predictably, precision improved as the number of species increased (Figure 2). This confirmed earlier findings of Boettiger et al. [-@Boettiger_et_al2012], and adhered to parametric statistical theory. However, in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: $\lambda_{est}= 0.0\to 0.985$), revealing that estimates of $\lambda$ were not a reliable indicator of input phylogenetic signal. Importantly, the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal. The worst precision was observed at intermediate levels of phylogenetic signal ($\lambda_{in}\approx0.5$), while precision improved as input levels approached the extremes of $\lambda$'s range (i.e., $\lambda_{in}\rightarrow0$ & $\lambda_{in}\rightarrow1$). Thus, estimates of $\lambda$ were least reflective of the true input signal at intermediate values. Additionally, even at large levels of species richness, we found that the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; $\lambda_{in} = 0.25 \to 0.65$). These findings were particularly unsettling when considered in light of our literature survey. Over one quarter of the $\lambda$ estimates published in empirical studies (421 of 1,572) were between $\lambda=0.25$ and $\lambda=0.75$ (Figure 1). This range reflected the region that our simulations identified as being the least reliable in terms of accurately characterizing levels of phylogenetic signal, yet 30% of these mid-range empirical estimates were explicitly interpreted in terms of the strength of phylogenetic signal that they represented (i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

When phylogenetic signal was investigated on balanced or pectinate trees, patterns in the precision of $\lambda$ were largely the same, with decreased precision at intermediate levels of phylogenetic signal (Supporting Information). Likewise, when $\lambda$ was co-estimated with regression parameters in PGLS regression and ANOVA, the results of our simulations were quite similar. Regression parameters ($\beta$) were accurately estimated, confirming earlier findings of Revell [-@Revell2010] (see Supporting Information). However, estimates of phylogenetic signal ($\lambda$) were less precise (Figure 3; see also Supporting Information), and the spread of $\lambda_{est}$ was similar to that observed when $\lambda$ was estimated for only the dependent variable, as in Figure 2. Taken together, these findings reveal that $\lambda_{est}$ does not precisely characterize known levels of phylogenetic signal in phenotypic datasets, and that biological interpretations of the strength of phylogenetic signal based on $\lambda$ may be highly inaccurate. \hfill\break

@@

An effect size based on $\lambda$ is a little more challenging because, as a statistic, it does not have a sampling distribution on which to measure standard error.  Confidence intervals for $\lambda$ are not generated from standard error calculations, but rather for the values of lambda that intersect the log-likehihood profile for corresponding percentiles of the $\chi^2$ distribution used to compare the compare the putative appropriate model to a null model with $\lambda = 0$ [citation needed, think Boettiger has a paper on this].  The simplest similar effect size is

\begin{align}
   \lvert Z_{\lambda} \rvert = \sqrt{\chi^2_{\hat{\lambda}}}
\end{align}

where, $\hat{\lambda}$ is the maximized likelihood value of $\lambda$ and $\chi^{2}_{\hat{\lambda}}$ is the likelihood ratio statistic for the value.  Hypothetically, this equation would be better represented as $Z_{\lambda} = d \sqrt{\chi^2_{\hat{\lambda}}}$, where $d$ is a binary value $(-1,1)$ to indicate a direction based on whether $\hat{\lambda}$ is below or above a critical value of $\lambda$, for a quantile from a $\chi^{2}$ distribution at a probability of 0.5.  The critical value would correspond to a value of $Z = 0$; i.e., the strength of phylogenetic signal that might be observed at the 50th percentile in a distribution of random outcomes, generated from a  model where $\lambda = 0$.  However, in some preliminary simulations, we found that mapping $\chi^2_{\hat{\lambda}}$ values this way did not produce effect sizes that were symmetrical about $Z = 0$, as the mapping is not linear and, especially for small trees, the log-likelihood profiles can be rather flat, making estimation of $\hat{\lambda}$ rather imprecise (as we elaborate more below.)  We will heneforth use equation 3, recognizing that large values of $Z$ would likely be positive, anyway, based on the reference model for the likelihood ratio test.

@@

To evaluate the utility of $Z_\kappa$ and $Z_\lambda$ we calculated both effect sizes for the simulated datasets generated above, and summarized the precision of each using its variance ($\sigma^2_{Z_\kappa}$ and $\sigma^2_{Z_\lambda}$, Figure 4: additional results in the Supporting Information). Here two things are evident. First, estimates of $Z_\kappa$ linearly track the input phylogenetic signal whereas estimates of $Z_\lambda$ do not (Figure 4A, B). Thus, actual changes in the strength of phylogenetic signal are reflected more evenly in the corresponding values of the effect size $Z_\kappa$. Second, the precision of $Z_\kappa$ is considerably more stable as compared with $Z_\lambda$. This may be seen by calculating the coefficients of variation for the set of precision estimates (i.e., $\sigma^2_{Z_\kappa}$ and $\sigma^2_{Z_\lambda}$) across input levels of phylogenetic signal.  Coefficients of variation in the precision of $Z_\kappa$ were up to an order of magnitude smaller than for $Z_\lambda$ (Figure 4C), implying that estimates of the strength of phylogenetic signal were more reliable and robust when using $Z_\kappa$. \hfill\break

@@

To demonstrate the utility of $\hat{Z}_{12}$ we quantified and compared the strength of phylogenetic signal of two phenotypic traits across species of plethodontid salamander. The data were part of a series of studies examining macroevolutionary trends in phenotypic diversification in this group [@Baken2019; @Baken2020]. Our dataset contained surface area to volume ratios (SA:V) and relative body width ($\frac{BW}{SVL}$) for 305 species (Figure 5A). For SA:V, 11 linear body measurements from 2,781 individuals were taken, from which estimates of the surface area and volume of the head, body, and tail were calculated and subsequently combined to arrive at the SA:V for each indiviual [for mathematical details see @Baken2020]. Species means were then obtained. Likewise, body size (SVL) and body width (BW) measurements were taken from 3,371 individuals, and species means of relative body width ($\frac{BW}{SVL}$) were calculated [data from @Baken2019]. A time-dated molecular phylogeny for the group [@Bonett2017] was then pruned to match the species in the phenotypic dataset. The phylogenetic signal in each trait was then characterized using $\kappa$, which was converted to its effect size ($Z_\kappa$) using `geomorph` 3.2.1 [@AdamsOtarola2013; @AdamsGeomorph]. Finally, the strength of phylogenetic signal was compared across traits using $\hat{Z}_{12}$ as described above (**to be incorporated in `geomorph` upon manuscript acceptance**). \hfill\break
